Machine Learning, Architectural Styles and Property Values

### 2019-09-14 text/manuscript_assa.Rmd
reorganised data section

---
title: "Machine Learning, Building Styles and Property Values"
author: "Thies Lindenthal* (University of Cambridge) & Erik B. Johnson (University of Alabama)"
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: Externalities of architecture, machine learning
fig_caption: yes
header-includes:
   - \usepackage{soul}
   - \usepackage{color}
   - \sethlcolor{yellow}
   - \usepackage{float}
   - \usepackage{booktabs}
   - \usepackage{graphicx}
   - \usepackage{grffile}
   - \usepackage{bm}
   - \usepackage{amsmath}
   - \usepackage{setspace}
   - \usepackage{multirow}
   - \doublespacing
   - \usepackage[backend=bibtex, style=authoryear]{biblatex}
fig_width: 6
fig_height: 4
output: 
  pdf_document:
    keep_tex: true
    number_sections: true
bibliography: references.bib
abstract: "This paper first introduces an algorithm that collects pictures of individual buildings from Google Street View. Second, it trains a deep convolutional neural network (CNN) to classify residential buildings into architectural styles, taking into account spatial dependencies of these styles. Third, it investigates whether architectural styles influence house prices. For re-sales, the architectural style is a significant determinant of transaction prices while no such effect is found for new buildings. Additionally, we are able to provide guidance on how to detect and overcome some of the limitations of machine learning methods through a large-scale comparison of predictions and expert classifications." 
---

@@

# Methodology

@@

Two final-year architectural students classified a large sub-sample of approximately 25,000 images from our data set of Cambridge houses. This is a much larger sample than needed in our case. In our case, each category requires less than 250 observations to reach saturated training accuracy levels. We greatly exceed this number for the purpose of this paper so that we can compare the out-of-sample convolutional neural network predictions to the architects' classifications. This allows us to examine the power and size of the assignment tests. In addition having both human and machine classification for a large sample of the data allows for robustness checks on the machine comparisons.

@@

* an input layer the size of the feature vectors,

@@

## Building style and property price

@@

Here, the natural logarithm of sales prices is explained by a linear combination of hedonic attributes described in vector $\mathbf{X}$, vectors of year $\mathbf{Y}$ and neighborhood $\bm{Loc}$ dummy variables and the building's estimated $\mathbf{Style}$ and the prevailing vintage of other buildings in the direct proximity $(\mathbf{StyleNeigh})$. The intercept is denoted by $\alpha$ while $\bm{\beta}$, $\bm{\delta}$, $\bm{\eta}$, $\bm{\iota}$, $\bm{\gamma}$ and $\bm{\lambda}$ are vectors of regression coefficients. $\epsilon$ is the IID error term. Heteroscedastic robust standard errors will be reported.

Are buildings with different appearances imperfect substitutes catering to multiple groups of households with distinct style preferences? @Buitelaar2017 indicate that any premium for an architectural style must stem from either differences in construction prices (which they do not find in their Dutch sample) or from supply constraints, as new construction potentially does not capture the demand for traditional styles. For Cambridge, new supply will inevitably be of either contemporary or revival style as historic vintages (and their styles), by definition, are not supplied any more. Estimating Eq. \ref{eq:hed} for a subset of newly constructed buildings will show whether construction prices or supply constraints for new homes built according to different architectural styles persist -- if too few vernacular buildings were built, prices should reflect such a shortage.

# Data

Residential real estate transactions are public data in the UK, collected and published by the Land Registry [@LandRegistry2016a]. The records include the date of transaction, price paid, street address, a classification of the property type (flat, detached, semi-detached, or terraced house), the estate type (freehold or leasehold) and an indicator for newly built properties. We select transactions from Cambridge proper which were recorded between January 1995 and October 2018, excluding any leaseholds, flats and properties classified as type "other", and sales with prices below £50,000 or in excess of £2,000,000. Linking recently taken images to sales data from 24 years assumes that houses have not changed their architectural styles -- which seems reasonable. We are not aware of any conversions of just the exterior of buildings. Full redevelopments did occur and we exclude sales for a given address pre-dating any redevelopment. Table \ref{tab:sumstats} presents summary statistics for the sample.

Overall, we collect pictures of more than 48,000 distinct buildings from the Google Street View API. The subset of 25,000 pictures that has been classified by trained architects into the seven styles will serve as \emph{ground truth} for our study. Two additional categories identify buildings hidden by *greenery* or uninformative pictures featuring *plain walls*. The classified images are available at the authors' websites.

@@

\begin{center}
  -- Insert Table \ref{tab:sumstats} and Table \ref{tab:hedreg} about here --
\end{center}

>>>>

---
title: "Machine Learning, Architectural Styles and Property Values"
author: "Thies Lindenthal* (University of Cambridge) & Erik B. Johnson (University of Alabama)"
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: Externalities of architecture, machine learning
fig_caption: yes
header-includes:
   - \usepackage{soul}
   - \usepackage{color}
   - \sethlcolor{yellow}
   - \usepackage{float}
   - \usepackage{booktabs}
   - \usepackage{graphicx}
   - \usepackage{grffile}
   - \usepackage{bm}
   - \usepackage{amsmath}
   - \usepackage{setspace}
   - \usepackage{multirow}
   - \doublespacing
   - \usepackage[backend=bibtex, style=authoryear]{biblatex}
fig_width: 6
fig_height: 4
output: 
  pdf_document:
    keep_tex: true
    number_sections: true
bibliography: references.bib
abstract: "This paper first introduces an algorithm that collects pictures of individual buildings from Google Street View. Second, it trains a deep convolutional neural network (CNN) to classify residential buildings into architectural styles, taking into account spatial dependencies of these styles. Third, it investigates whether architectural styles influence house prices. For re-sales, the architectural style is a significant determinant of transaction prices while no such effect is found for new buildings. Additionally, we are able to provide guidance on how to detect and overcome some of the limitations of machine learning methods through a large-scale comparison of predictions and expert classifications." 
---

@@

# ML classification of building styles

@@

Two final-year architectural students classified a large sub-sample of approximately 25,000 images from our data set of Cambridge houses.\footnote{The full set of classified images are available for download at the authors' websites.} This is a much larger sample than needed in our case. In our case, each category requires less than 250 observations to reach saturated training accuracy levels. We greatly exceed this number for the purpose of this paper so that we can compare the out-of-sample convolutional neural network predictions to the architects' classifications. This allows us to examine the power and size of the assignment tests. In addition having both human and machine classification for a large sample of the data allows for robustness checks on the machine comparisons.

@@

* an input layer the size of the feature vectors, e.g. 2048 or 4096, respectively,

@@

# Architectural styles and property values

## Residential transaction data

Residential real estate transactions are public data in the UK, collected and published by the Land Registry [@LandRegistry2016a]. The records include the date of transaction, price paid, street address, a classification of the property type (flat, detached, semi-detached, or terraced house), the estate type (freehold or leasehold) and an indicator for newly built properties. We select transactions from Cambridge proper which were recorded between January 1995 and October 2018, excluding any leaseholds, flats and properties classified as type "other", and sales with prices below £50,000 or in excess of £2,000,000. Linking recently taken images to sales data from 24 years assumes that houses have not changed their architectural styles -- which seems reasonable. We are not aware of any conversions of just the exterior of buildings. In case of full redevelopments, we exclude any sales for a given address pre-dating the redevelopment, ensuring that a currently taken pictures represents the house at all included transactions. Table \ref{tab:sumstats} presents summary statistics for the sample.

\begin{center}
  -- Insert Table \ref{tab:sumstats} about here --
\end{center}

## Hedonic regression setup

@@

Here, the natural logarithm of sales prices is explained by a linear combination of hedonic attributes described in vector $\mathbf{X}$, vectors of year $\mathbf{Y}$ and neighborhood $\bm{Loc}$ dummy variables and the building's estimated $\mathbf{Style}$ and the prevailing styles of other buildings in the direct proximity $(\mathbf{StyleNeigh})$.  $\mathbf{Style\cdot StyleNeigh}$ is a vector of interaction terms for the building's and the  neighborhood's dominant style. The intercept is denoted by $\alpha$ while $\bm{\beta}$, $\bm{\delta}$, $\bm{\eta}$, $\bm{\iota}$, $\bm{\gamma}$ and $\bm{\lambda}$ are vectors of regression coefficients. $\epsilon$ is the IID error term. Heteroscedastic robust standard errors will be reported.

Are buildings with different appearances imperfect substitutes catering to multiple groups of households with distinct style or vintage preferences? @Buitelaar2017 indicate that any premium for an architectural style must stem from either differences in construction prices (which they do not find in their Dutch sample) or from supply constraints, as new construction potentially does not capture the demand for traditional styles. For Cambridge, new supply will inevitably be of either contemporary or revival style as historic vintages (and their styles), by definition, are not supplied any more. Estimating Eq. \ref{eq:hed} for a subset of newly constructed buildings will show whether construction prices or supply constraints for new homes built according to different architectural styles persist -- if too few vernacular buildings were built, prices should reflect such a shortage. Singling out new buildings crucially controls for the otherwise unobserved age of buildings, which is tightly intertwined with its aesthetics.

@@

\begin{center}
  -- Insert Table \ref{tab:hedreg} about here --
\end{center}

### 2019-09-14 text/manuscript_assa.tex
reorganised data section

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Machine Learning, Building Styles and Property Values},
            pdfauthor={Thies Lindenthal* (University of Cambridge) \& Erik B. Johnson (University of Alabama)},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Machine Learning, Building Styles and Property Values}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Thies Lindenthal* (University of Cambridge) \& Erik B. Johnson
(University of Alabama)}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{13 September 2019}

@@

\hypertarget{methodology}{%
\section{Methodology}\label{methodology}}

@@

\begin{itemize}
\item
  an input layer the size of the feature vectors,
\item
  one dense layer (relu) half the size of the input layer, one
  subsequent dropout (rate 0.5) layer,
\item
  one dense layer (relu) a quarter the size of the input layer, one
  subsequent dropout (rate 0.5) layer,
\item
  and the final dense output layer with softmax activation.
\end{itemize}

@@

\hypertarget{building-style-and-property-price}{%
\subsection{Building style and property
price}\label{building-style-and-property-price}}

@@

Here, the natural logarithm of sales prices is explained by a linear
combination of hedonic attributes described in vector \(\mathbf{X}\),
vectors of year \(\mathbf{Y}\) and neighborhood \(\bm{Loc}\) dummy
variables and the building's estimated \(\mathbf{Style}\) and the
prevailing vintage of other buildings in the direct proximity
\((\mathbf{StyleNeigh})\). The intercept is denoted by \(\alpha\) while
\(\bm{\beta}\), \(\bm{\delta}\), \(\bm{\eta}\), \(\bm{\iota}\),
\(\bm{\gamma}\) and \(\bm{\lambda}\) are vectors of regression
coefficients. \(\epsilon\) is the IID error term. Heteroscedastic robust
standard errors will be reported.

Are buildings with different appearances imperfect substitutes catering
to multiple groups of households with distinct style preferences?
Buitelaar and Schilder (2017) indicate that any premium for an
architectural style must stem from either differences in construction
prices (which they do not find in their Dutch sample) or from supply
constraints, as new construction potentially does not capture the demand
for traditional styles. For Cambridge, new supply will inevitably be of
either contemporary or revival style as historic vintages (and their
styles), by definition, are not supplied any more. Estimating Eq.
\ref{eq:hed} for a subset of newly constructed buildings will show
whether construction prices or supply constraints for new homes built
according to different architectural styles persist -- if too few
vernacular buildings were built, prices should reflect such a shortage.

\hypertarget{data}{%
\section{Data}\label{data}}

Residential real estate transactions are public data in the UK,
collected and published by the Land Registry (Land Registry 2017). The
records include the date of transaction, price paid, street address, a
classification of the property type (flat, detached, semi-detached, or
terraced house), the estate type (freehold or leasehold) and an
indicator for newly built properties. We select transactions from
Cambridge proper which were recorded between January 1995 and October
2018, excluding any leaseholds, flats and properties classified as type
``other'', and sales with prices below £50,000 or in excess of
£2,000,000. Linking recently taken images to sales data from 24 years
assumes that houses have not changed their architectural styles -- which
seems reasonable. We are not aware of any conversions of just the
exterior of buildings. Full redevelopments did occur and we exclude
sales for a given address pre-dating any redevelopment. Table
\ref{tab:sumstats} presents summary statistics for the sample.

Overall, we collect pictures of more than 48,000 distinct buildings from
the Google Street View API. The subset of 25,000 pictures that has been
classified by trained architects into the seven styles will serve as
\emph{ground truth} for our study. Two additional categories identify
buildings hidden by \emph{greenery} or uninformative pictures featuring
\emph{plain walls}. The classified images are available at the authors'
websites.

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Machine Learning, Architectural Styles and Property Values},
            pdfauthor={Thies Lindenthal* (University of Cambridge) \& Erik B. Johnson (University of Alabama)},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Machine Learning, Architectural Styles and Property Values}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Thies Lindenthal* (University of Cambridge) \& Erik B. Johnson
(University of Alabama)}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{14 September 2019}

@@

\hypertarget{ml-classification-of-building-styles}{%
\section{ML classification of building
styles}\label{ml-classification-of-building-styles}}

@@

\begin{itemize}
\item
  an input layer the size of the feature vectors, e.g.~2048 or 4096,
  respectively,
\item
  one dense layer (relu) half the size of the input layer, one
  subsequent dropout (rate 0.5) layer,
\item
  one dense layer (relu) a quarter the size of the input layer, one
  subsequent dropout (rate 0.5) layer,
\item
  and the final dense output layer with softmax activation.
\end{itemize}

@@

\hypertarget{architectural-styles-and-property-values}{%
\section{Architectural styles and property
values}\label{architectural-styles-and-property-values}}

\hypertarget{residential-transaction-data}{%
\subsection{Residential transaction
data}\label{residential-transaction-data}}

Residential real estate transactions are public data in the UK,
collected and published by the Land Registry (Land Registry 2017). The
records include the date of transaction, price paid, street address, a
classification of the property type (flat, detached, semi-detached, or
terraced house), the estate type (freehold or leasehold) and an
indicator for newly built properties. We select transactions from
Cambridge proper which were recorded between January 1995 and October
2018, excluding any leaseholds, flats and properties classified as type
``other'', and sales with prices below £50,000 or in excess of
£2,000,000. Linking recently taken images to sales data from 24 years
assumes that houses have not changed their architectural styles -- which
seems reasonable. We are not aware of any conversions of just the
exterior of buildings. In case of full redevelopments, we exclude any
sales for a given address pre-dating the redevelopment, ensuring that a
currently taken pictures represents the house at all included
transactions. Table \ref{tab:sumstats} presents summary statistics for
the sample.

\begin{center}
  -- Insert Table \ref{tab:sumstats} about here --
\end{center}

\hypertarget{hedonic-regression-setup}{%
\subsection{Hedonic regression setup}\label{hedonic-regression-setup}}

@@

Here, the natural logarithm of sales prices is explained by a linear
combination of hedonic attributes described in vector \(\mathbf{X}\),
vectors of year \(\mathbf{Y}\) and neighborhood \(\bm{Loc}\) dummy
variables and the building's estimated \(\mathbf{Style}\) and the
prevailing styles of other buildings in the direct proximity
\((\mathbf{StyleNeigh})\). \(\mathbf{Style\cdot StyleNeigh}\) is a
vector of interaction terms for the building's and the neighborhood's
dominant style. The intercept is denoted by \(\alpha\) while
\(\bm{\beta}\), \(\bm{\delta}\), \(\bm{\eta}\), \(\bm{\iota}\),
\(\bm{\gamma}\) and \(\bm{\lambda}\) are vectors of regression
coefficients. \(\epsilon\) is the IID error term. Heteroscedastic robust
standard errors will be reported.

Are buildings with different appearances imperfect substitutes catering
to multiple groups of households with distinct style or vintage
preferences? Buitelaar and Schilder (2017) indicate that any premium for
an architectural style must stem from either differences in construction
prices (which they do not find in their Dutch sample) or from supply
constraints, as new construction potentially does not capture the demand
for traditional styles. For Cambridge, new supply will inevitably be of
either contemporary or revival style as historic vintages (and their
styles), by definition, are not supplied any more. Estimating Eq.
\ref{eq:hed} for a subset of newly constructed buildings will show
whether construction prices or supply constraints for new homes built
according to different architectural styles persist -- if too few
vernacular buildings were built, prices should reflect such a shortage.
Singling out new buildings crucially controls for the otherwise
unobserved age of buildings, which is tightly intertwined with its
aesthetics.
